---
title: "analysis"
output: html_document
date: "2026-05-11"
---

```{r}
library(tidyverse)
library(caret)
library(tidymodels)
library(glmnet)
library(earth)
library(gam)
library(pROC)
library(pdp)
library(vip)
library(gridExtra)
library(gbm)
```

### Loading and Cleaning

```{r Cleaning}
load("severity.RData")

# Dropping rows with missing values and ID variable
dat <- na.omit(dat) %>%
  select(-id)

# Factoring & Labeling

dat$gender <- factor(dat$gender, levels = c(0, 1), labels = c("Female", "Male"))
dat$vaccine <- factor(dat$vaccine, levels = c(0, 1), labels = c("Unvaccinated", "Vaccinated"))
dat$hypertension <- factor(dat$hypertension, levels = c(0, 1), labels = c("No", "Yes"))
dat$diabetes <- factor(dat$diabetes, levels = c(0, 1), labels = c("No", "Yes"))
dat$severity <- factor(dat$severity, levels = c(0, 1), labels = c("not_severe", "severe"))
dat$race <- factor(dat$race, levels = c(1, 2, 3, 4), labels = c("White", "Asian", "Black", "Hispanic"))
dat$smoking <- factor(dat$smoking, levels = c(0, 1, 2), labels = c("Never", "Former", "Current"))
```

### Data preperation and partitioning for modeling

```{r}
# Random 1000 sample with Uni
set.seed(3015)
my_dat <- dat[sample(nrow(dat), 1000), ]

# 80/20 split
set.seed(3015)
data_split <- initial_split(my_dat, prop = 0.80)
training_data <- training(data_split)
testing_data <- testing(data_split)
```

### Summary Statistics

```{r}
summary(my_dat)

# Combined Exploratory Figure
featurePlot(x = my_dat[, c("age", "bmi", "SBP", "LDL", "depression")], 
            y = my_dat$severity, 
            plot = "box", 
            scales = list(x = list(relation = "free"), y = list(relation = "free")),
            main = "Distribution of Clinical Factors by Severity")
```

---

# Training Simpler Models 

### Logistic Regression
```{r}
# CV Control
ctrl <- trainControl(method = "cv", number = 10, 
                     summaryFunction = twoClassSummary, 
                     classProbs = TRUE)

# Baseline: Simple Logistic Regression (GLM) 
set.seed(3015)
model.glm <- train(severity ~ ., data = training_data,
                   method = "glm",
                   metric = "ROC",
                   trControl = ctrl)

summary(model.glm$finalModel)
```

### Penalized Logistic Regression (Elastic Net) 

```{r}
glmnGrid <- expand.grid(.alpha = seq(0, 1, length = 21),
                        .lambda = exp(seq(-8, -1, length = 50)))
set.seed(3015)
model.enet <- train(severity ~ ., data = training_data,
                    method = "glmnet",
                    tuneGrid = glmnGrid,
                    metric = "ROC",
                    trControl = ctrl)

model.enet$bestTune

myCol = rainbow(25)
myPar = list(superpose.symbol = list(col = myCol),
             superpose.line = list(col = myCol))

plot(model.enet , par.settings = myPar, xTrans = log)
```

### Multivariate Adaptive Regression Splines (MARS)

```{r}
set.seed(3015)
model.mars <- train(severity ~ ., 
                    data = training_data,
                    method = "earth", 
                    tuneGrid = expand.grid(degree = 1:4, nprune = 2:20),
                    metric = "ROC",
                    trControl = ctrl)

ggplot(model.mars)
model.mars$bestTune
```

### Generalized Additive Model (GAM)

```{r}
set.seed(3015)
model.gam <- train(severity ~ .,
                   data = training_data,
                   method = "gam", 
                   metric = "ROC", 
                   trControl = ctrl)

model.gam$bestTune
model.gam$finalModel
plot(model.gam$finalModel, se = TRUE)
summary(model.gam$finalModel)

par(mfrow = c(2, 2)) 


plot(model.gam$finalModel, 
     shade = TRUE, 
     seWithMean = TRUE, 
     main = "Clinical Nonlinear Trends (2021-2023)", 
     col = "blue", 
     pch = 16)

par(mfrow = c(1, 1))
```

# Training Advanced models

```{r}
# Boosting (GBM) 
gbmA.grid <- expand.grid(n.trees = c(100, 500, 1000, 2000, 5000), 
                        interaction.depth = 1:5, 
                        shrinkage = c(0.001, 0.003, 0.005),
                        n.minobsinnode = 10) 
set.seed(3015)
model.gbm <- train(severity ~ ., 
                   data = training_data, 
                   method = "gbm",
                   distribution = "adaboost",
                   tuneGrid = gbmA.grid,
                   metric = "ROC",
                   trControl = ctrl, 
                   verbose = FALSE) 

ggplot(model.gbm, highlight = TRUE)
```


```{r}
# SVM with Radial Kernel 
svmr.grid <- expand.grid(C = exp(seq(1, 7, len = 20)), 
                         sigma = exp(seq(-10, -2, len = 10))) 
set.seed(3015)
model.svm <- train(severity ~ ., 
                    data = training_data, 
                    method = "svmRadialSigma",
                    tuneGrid = svmr.grid, 
                    metric = "ROC", 
                    trControl = ctrl) 
myCol <- rainbow(25)
myPar <- list(superpose.symbol = list(col = myCol),
superpose.line = list(col = myCol))
plot(model.svm, highlight = TRUE, par.settings = myPar)
```

### Tuning visual for all models with grids

```{r}
# Tuning plots 
# Elastic Net tuning plot 
p_enet <- plot(model.enet, par.settings = myPar, xTrans = log, main = "Enet Tuning")

# MARS tuning plot 
p_mars <- ggplot(model.mars) + ggtitle("MARS Tuning")

# Boosting tuning plot 
p_gbm <- ggplot(model.gbm, highlight = TRUE) + ggtitle("Boosting Tuning")

# SVM tuning plot 
p_svm <- plot(model.svm, highlight = TRUE, par.settings = myPar, main = "SVM Tuning")

# Grid
grid.arrange(p_enet, p_mars, p_gbm, p_svm, 
             nrow = 2, ncol = 2,
             top = "Figure: Model Selection and Hyperparameter Tuning")
```


```{r}
# Variable Importance Plot for the Advanced Models
vip(model.gbm, method = "model") + ggtitle("Key Factors for COVID-19 Severity (Boosting)")

# Partial Dependence Plots for most important continuous variables
# p1 - Age
p1 <- pdp::partial(model.gbm, pred.var = "age", grid.resolution = 100, prob = TRUE) %>%
  autoplot(rug = TRUE, train = training_data) + 
  ggtitle("Influence of Age") +
  ylab("Prob(Severe)")

# p2 -  Systolic Blood Pressure (SBP)
p2 <- pdp::partial(model.gbm, pred.var = "SBP", grid.resolution = 100, prob = TRUE) %>%
  autoplot(rug = TRUE, train = training_data) + 
  ggtitle("Influence of SBP") +
  ylab("Prob(Severe)")

# p3 - Body Mass Index (BMI)
p3 <- pdp::partial(model.gbm, pred.var = "bmi", grid.resolution = 100, prob = TRUE) %>%
  autoplot(rug = TRUE, train = training_data) + 
  ggtitle("Influence of BMI") +
  ylab("Prob(Severe)")

# Combining
grid.arrange(p1, p2, p3, nrow = 1)
```

---

# Internal Model Comparison
```{r}
# Compare all trained models 
resamp <- resamples(list(
  GLM = model.glm, 
  ENET = model.enet, 
  MARS = model.mars, 
  GAM = model.gam, 
  Boosting = model.gbm, 
  SVM = model.svm
)) 

# Summarize the performance metrics (ROC, Sens, Spec) across resamples
summary(resamp) 

# Visualize the ROC performance comparison
bwplot(resamp, metric = "ROC") 
```

---

# Model Evaluation of top performers on Test Set

```{r}
# Final validation on the Test Set (n=200) of best simple and advanced models
probs_glm  <- predict(model.glm,  newdata = testing_data, type = "prob")[, "severe"]
probs_enet <- predict(model.enet, newdata = testing_data, type = "prob")[, "severe"]
probs_gbm  <- predict(model.gbm,  newdata = testing_data, type = "prob")[, "severe"]
probs_svm  <- predict(model.svm,  newdata = testing_data, type = "prob")[, "severe"]

# Final Test ROC
roc_glm_final  <- roc(testing_data$severity, probs_glm)
roc_enet_final <- roc(testing_data$severity, probs_enet) 
roc_gbm_final  <- roc(testing_data$severity, probs_gbm)
roc_svm_final  <- roc(testing_data$severity, probs_svm)

# Create the final comparison figure
ggroc(list(GLM = roc_glm_final, 
           ENET = roc_enet_final, 
           Boosting = roc_gbm_final, 
           SVM = roc_svm_final), 
      legacy.axes = TRUE) +
  geom_abline(intercept = 0, slope = 1, color = "grey") +
  ggtitle("Final Test Set Performance Comparison")


auc_table <- data.frame(
  Model = c("Baseline GLM", "Elastic Net (Best Simple)", "Boosting", "SVM"),
  Test_AUC = c(auc(roc_glm_final), auc(roc_enet_final), auc(roc_gbm_final), auc(roc_svm_final))
)
print(auc_table)
```

---

# Predictive Risk Score

```{r}
# Generate the risk scores (probabilities) using the best-performing model (Boosting)
test_risk_scores <- predict(model.gbm, newdata = testing_data, type = "prob")[, "severe"]

# Attach scores to a summary table to demonstrate the clinical tool
risk_score_report <- testing_data %>%
  mutate(Risk_Score = round(test_risk_scores, 4)) %>%
  select(age, gender, vaccine, bmi, Risk_Score)

# Display the first 10 participants to show the individual risk quantification
head(risk_score_report, 10)
```

