---
title: "analysis"
output: html_document
date: "2026-05-11"
---

```{r}
library(tidyverse)
library(knitr)
library(kableExtra)
library(caret)
library(tidymodels)
library(glmnet)
library(earth)
library(gam)
library(pROC)
library(pdp)
library(vip)
library(gridExtra)
library(gbm)
library(rpart.plot)
library(gtsummary)
```

### Loading and Cleaning

```{r Cleaning}
load("severity.RData")

# Dropping rows with missing values and ID variable
dat <- na.omit(dat) %>%
  select(-id)

# Factoring & Labeling

dat$gender <- factor(dat$gender, levels = c(0, 1), labels = c("Female", "Male"))
dat$vaccine <- factor(dat$vaccine, levels = c(0, 1), labels = c("Unvaccinated", "Vaccinated"))
dat$hypertension <- factor(dat$hypertension, levels = c(0, 1), labels = c("No", "Yes"))
dat$diabetes <- factor(dat$diabetes, levels = c(0, 1), labels = c("No", "Yes"))
dat$severity <- factor(dat$severity, levels = c(0, 1), labels = c("not_severe", "severe"))
dat$race <- factor(dat$race, levels = c(1, 2, 3, 4), labels = c("White", "Asian", "Black", "Hispanic"))
dat$smoking <- factor(dat$smoking, levels = c(0, 1, 2), labels = c("Never", "Former", "Current"))
```

### Data preperation and partitioning for modeling

```{r}
# Random 1000 sample with Uni
set.seed(3015)
my_dat <- dat[sample(nrow(dat), 1000), ]

# 80/20 split
set.seed(3015)
data_split <- initial_split(my_dat, prop = 0.80)
training_data <- training(data_split)
testing_data <- testing(data_split)
```

### Summary Statistics

```{r}
summary(my_dat)

my_dat %>%
  tbl_summary(
    by = severity, 
    statistic = list(all_continuous() ~ "{mean} ({sd})", 
                     all_categorical() ~ "{n} ({p}%)"),
    digits = all_continuous() ~ 1,
    label = list(
      age ~ "Age (years)",
      gender ~ "Gender",
      bmi ~ "Body Mass Index",
      vaccine ~ "Vaccination Status",
      SBP ~ "Systolic Blood Pressure",
      LDL ~ "LDL Cholesterol"
    )
  ) %>%
  add_p() %>% 
  modify_header(label = "**Variable**") %>%
  bold_labels()

# Combined Exploratory Figure
featurePlot(x = my_dat[, c("age", "bmi", "SBP", "LDL", "depression")], 
            y = my_dat$severity, 
            plot = "box", 
            scales = list(x = list(relation = "free"), y = list(relation = "free")),
            main = "Distribution of Clinical Factors by Severity")
```

---

# Training Simpler Models 

### Logistic Regression
```{r}
# CV Control
ctrl <- trainControl(method = "cv", number = 10, 
                     summaryFunction = twoClassSummary, 
                     classProbs = TRUE)

# Baseline: Simple Logistic Regression (GLM) 
set.seed(3015)
model.glm <- train(severity ~ ., data = training_data,
                   method = "glm",
                   metric = "ROC",
                   trControl = ctrl)

summary(model.glm$finalModel)
```

### Penalized Logistic Regression (Elastic Net) 

```{r}
glmnGrid <- expand.grid(.alpha = seq(0, 1, length = 21),
                        .lambda = exp(seq(-8, -1, length = 50)))
set.seed(3015)
model.enet <- train(severity ~ ., data = training_data,
                    method = "glmnet",
                    tuneGrid = glmnGrid,
                    metric = "ROC",
                    trControl = ctrl)

model.enet$bestTune

myCol = rainbow(25)
myPar = list(superpose.symbol = list(col = myCol),
             superpose.line = list(col = myCol))

plot(model.enet , par.settings = myPar, xTrans = log)
```

### Multivariate Adaptive Regression Splines (MARS)

```{r}
set.seed(3015)
model.mars <- train(severity ~ ., 
                    data = training_data,
                    method = "earth", 
                    tuneGrid = expand.grid(degree = 1:4, nprune = 2:20),
                    metric = "ROC",
                    trControl = ctrl)

ggplot(model.mars)
model.mars$bestTune
```

### Generalized Additive Model (GAM)

```{r}
set.seed(3015)
model.gam <- train(severity ~ .,
                   data = training_data,
                   method = "gam", 
                   metric = "ROC", 
                   trControl = ctrl)

model.gam$bestTune
model.gam$finalModel
plot(model.gam$finalModel, se = TRUE)
summary(model.gam$finalModel)

par(mfrow = c(2, 2)) 


plot(model.gam$finalModel, 
     shade = TRUE, 
     seWithMean = TRUE, 
     main = "Clinical Nonlinear Trends (2021-2023)", 
     col = "blue", 
     pch = 16)

par(mfrow = c(1, 1))
```

### Classification Tree

```{r}
grid.tree <- expand.grid(cp = seq(0.0001, 0.05, length = 50))

set.seed(3015) # Using your consistent project seed
model.tree <- train(severity ~ ., 
                   data = training_data, 
                   method = "rpart", 
                   metric = "ROC", 
                   tuneGrid = grid.tree, 
                   trControl = ctrl)

model.tree

plot(model.tree)

rpart.plot(model.tree$finalModel, type = 4, extra = 101, under = TRUE, cex = 0.8,
           main = "Decision Tree for COVID-19 Severity")
```


# Training Advanced models

```{r}
# Boosting (GBM) 
gbmA.grid <- expand.grid(n.trees = c(100, 500, 1000, 2000, 5000), 
                        interaction.depth = 1:5, 
                        shrinkage = c(0.001, 0.003, 0.005),
                        n.minobsinnode = 10) 
set.seed(3015)
model.gbm <- train(severity ~ ., 
                   data = training_data, 
                   method = "gbm",
                   distribution = "adaboost",
                   tuneGrid = gbmA.grid,
                   metric = "ROC",
                   trControl = ctrl, 
                   verbose = FALSE) 

ggplot(model.gbm, highlight = TRUE)
```


```{r}
# SVM with Radial Kernel 
svmr.grid <- expand.grid(C = exp(seq(1, 7, len = 20)), 
                         sigma = exp(seq(-10, -2, len = 10))) 
set.seed(3015)
model.svm <- train(severity ~ ., 
                    data = training_data, 
                    method = "svmRadialSigma",
                    tuneGrid = svmr.grid, 
                    metric = "ROC", 
                    trControl = ctrl) 
myCol <- rainbow(25)
myPar <- list(superpose.symbol = list(col = myCol),
superpose.line = list(col = myCol))
plot(model.svm, highlight = TRUE, par.settings = myPar)
```

### Tuning visual for all models with grids

```{r}
# Tuning plots 
# Elastic Net tuning plot 
p_enet <- plot(model.enet, par.settings = myPar, xTrans = log, main = "Enet Tuning")

# MARS tuning plot 
p_mars <- ggplot(model.mars) + ggtitle("MARS Tuning")

# Classification Tree
p_tree <- ggplot(model.tree, highlight = TRUE) + ggtitle("Decision Tree Tuning")

# Boosting tuning plot 
p_gbm <- ggplot(model.gbm, highlight = TRUE) + ggtitle("Boosting Tuning")

# SVM tuning plot 
p_svm <- plot(model.svm, highlight = TRUE, par.settings = myPar, main = "SVM Tuning")

# Grid
grid.arrange(p_enet, p_mars, p_tree, p_gbm, p_svm, 
             nrow = 3, ncol = 2,
             top = "Figure: Model Selection and Hyperparameter Tuning")
```


```{r}
# Variable Importance Plot for the Advanced Models
vip(model.gbm, method = "model") + ggtitle("Key Factors for COVID-19 Severity (Boosting)")

# Partial Dependence Plots for most important continuous variables
# p1 - Age
p1 <- pdp::partial(model.gbm, pred.var = "age", grid.resolution = 100, prob = TRUE) %>%
  autoplot(rug = TRUE, train = training_data) + 
  ggtitle("Influence of Age") +
  ylab("Prob(Severe)")

# p2 -  Systolic Blood Pressure (SBP)
p2 <- pdp::partial(model.gbm, pred.var = "SBP", grid.resolution = 100, prob = TRUE) %>%
  autoplot(rug = TRUE, train = training_data) + 
  ggtitle("Influence of SBP") +
  ylab("Prob(Severe)")

# p3 - Body Mass Index (BMI)
p3 <- pdp::partial(model.gbm, pred.var = "bmi", grid.resolution = 100, prob = TRUE) %>%
  autoplot(rug = TRUE, train = training_data) + 
  ggtitle("Influence of BMI") +
  ylab("Prob(Severe)")

# p4 - Vaccination status (binary)
p_vac <- pdp::partial(model.gbm, 
                      pred.var = "vaccine", 
                      which.class = 2, # Tells R to plot the 'severe' class
                      prob = TRUE) %>%
  ggplot(aes(x = vaccine, y = yhat, fill = vaccine)) +
  geom_col(width = 0.5) +
  theme_minimal() +
  labs(title = "Marginal Effect of Vaccination on COVID-19 Severity",
       y = "Predicted Probability of Severe Infection",
       x = "Vaccination Status") +
  scale_fill_manual(values = c("Unvaccinated" = "firebrick", "Vaccinated" = "forestgreen"))

# Combining
grid.arrange(p1, p2, p3, p_vac, nrow = 1)
```

---

# Internal Model Comparison
```{r}
# Compare all trained models 
resamp <- resamples(list(
  GLM = model.glm, 
  ENET = model.enet, 
  MARS = model.mars, 
  TREE = model.tree,
  GAM = model.gam, 
  Boosting = model.gbm, 
  SVM = model.svm
)) 

# Summarize the performance metrics (ROC, Sens, Spec) across resamples
summary(resamp) 

# Visualize the ROC performance comparison
bwplot(resamp, metric = "ROC") 
```

---

# Model Evaluation of top performers on Test Set

```{r}
# Final validation on the Test Set for ALL models
probs_glm   <- predict(model.glm,   newdata = testing_data, type = "prob")[, "severe"]
probs_enet  <- predict(model.enet,  newdata = testing_data, type = "prob")[, "severe"]
probs_mars  <- predict(model.mars,  newdata = testing_data, type = "prob")[, "severe"]
probs_tree  <- predict(model.tree,  newdata = testing_data, type = "prob")[, "severe"]
probs_gam   <- predict(model.gam,   newdata = testing_data, type = "prob")[, "severe"]
probs_gbm   <- predict(model.gbm,   newdata = testing_data, type = "prob")[, "severe"]
probs_svm   <- predict(model.svm,   newdata = testing_data, type = "prob")[, "severe"]

roc_glm_final   <- roc(testing_data$severity, probs_glm)
roc_enet_final  <- roc(testing_data$severity, probs_enet)
roc_mars_final  <- roc(testing_data$severity, probs_mars)
roc_tree_final  <- roc(testing_data$severity, probs_tree)
roc_gam_final   <- roc(testing_data$severity, probs_gam)
roc_gbm_final   <- roc(testing_data$severity, probs_gbm)
roc_svm_final   <- roc(testing_data$severity, probs_svm)

# Comparison Plot
roc_list_final <- list(
  GLM = roc_glm_final, 
  ENET = roc_enet_final, 
  MARS = roc_mars_final,
  Tree = roc_tree_final,
  GAM = roc_gam_final, 
  Boosting = roc_gbm_final, 
  SVM = roc_svm_final
)

p_final_roc <- ggroc(roc_list_final, legacy.axes = TRUE) +
  geom_abline(intercept = 0, slope = 1, color = "grey", linetype = "dashed") +
  theme_minimal() +
  labs(title = "Final Test Set Performance Comparison (All Models)",
       x = "1 - Specificity (False Positive Rate)",
       y = "Sensitivity (True Positive Rate)") +
  scale_color_discrete(name = "Model")


p_final_roc

# Final AUC values
get_roc_stats <- function(roc_obj) {
  stats <- coords(roc_obj, "best", ret = c("sensitivity", "specificity"), transpose = FALSE)
  return(stats[1, ]) 
}

auc_table_detailed <- data.frame(
  Model = c("Baseline GLM", "Elastic Net", "MARS", "Decision Tree", "GAM", "Boosting", "SVM"),
  Test_AUC = c(as.numeric(auc(roc_glm_final)), 
               as.numeric(auc(roc_enet_final)), 
               as.numeric(auc(roc_mars_final)), 
               as.numeric(auc(roc_tree_final)),
               as.numeric(auc(roc_gam_final)), 
               as.numeric(auc(roc_gbm_final)), 
               as.numeric(auc(roc_svm_final)))
)

roc_list <- list(roc_glm_final, roc_enet_final, roc_mars_final, 
                 roc_tree_final, roc_gam_final, roc_gbm_final, roc_svm_final)

metrics_df <- do.call(rbind, lapply(roc_list, get_roc_stats))
auc_table_detailed <- cbind(auc_table_detailed, metrics_df)

auc_table_detailed %>%
  arrange(desc(Test_AUC)) %>%
  kable(caption = "Table 3: Final Model Performance on Unseen Test Set (n=200)",
        col.names = c("Model", "Test AUC", "Sensitivity", "Specificity"),
        digits = 4, 
        booktabs = TRUE,
        align = "lccc") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F) %>%
  row_spec(1, bold = TRUE, background = "#DFF0D8") 

  auc_table_detailed
```

---

# Predictive Risk Score

```{r}
# Generate the risk scores (probabilities) using the best-performing model (MARS)
test_risk_scores <- predict(model.mars, newdata = testing_data, type = "prob")[, "severe"]

risk_score_report <- testing_data %>%
  mutate(Risk_Score = round(test_risk_scores, 4)) %>%
  select(age, gender, vaccine, bmi, Risk_Score)

# Display the first 10 participants to show example of the individual risk quantification
head(risk_score_report, 10)

# Creating a table to show the predictions by age groupings and vaccination status
risk_summary_data <- risk_score_report %>%
  mutate(
    age_group = cut(age, breaks = c(48, 65, 100), 
                    labels = c( "Middle-Aged (48-65)", "Senior (65+)")),
    Risk_Percent = Risk_Score * 100 # Convert to percentage for readability
  )

#  tbl_summary
risk_summary_table <- risk_summary_data %>%
  select(age_group, vaccine, gender, Risk_Percent) %>%
  tbl_summary(
    by = vaccine,
    include = c(age_group, gender, Risk_Percent),
    statistic = list(Risk_Percent ~ "{mean}% ({min}%, {max}%)"),
    label = list(
      age_group ~ "Age Category",
      gender ~ "Gender",
      Risk_Percent ~ "Predicted Risk of Severity (Mean, Range)"
    )
  ) %>%
  modify_header(label = "**Patient Characteristics**") %>%
  modify_spanning_header(all_stat_cols() ~ "**Predicted Risk by Vaccination Status**") %>%
  bold_labels()


risk_summary_table
```

